---
title: "R Notebook"
output: html_document
editor_options: 
  chunk_output_type: console
---
## Setup
```{r setup}
library(conflicted)
library(knitr)
library(tidyverse)
library(ape)
library(vegan)
library(ggrepel)
library(agricolae)
library(ggbeeswarm)
library(patchwork)


conflict_prefer("strsplit", "Biostrings")
conflict_prefer("filter", "dplyr")

options(max.print="999")
opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, prompt=FALSE, tidy = TRUE, comment = NA, message = FALSE)
opts_knit$set(width=75)
source("assets/utils.R")

# set path
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(rprojroot::find_rstudio_root_file())

path_data = "D:/Dropbox/Academia/Data/sscap_uga/"
sessionInfo()
```

## Set_up the analysis (No need to run)
### Some misc data files
```{r}
GG_taxa <- read.delim(file = paste0(path_data, "assets/gg_13_5_taxonomy.txt"), header = F, stringsAsFactors = F)
colnames(GG_taxa) <- c("OTU_ID", "Taxa")
```

### Read 16S enriched count file
```{r eval=FALSE}
Enriched150 <- read.table("16Scap/Enriched_PE150_OTU_count.txt", header = F, sep = "\t", stringsAsFactors = F) # a table with OTU id, count, sampleid, and library type
tmp <- Enriched150
colnames(tmp) <- c("OTU_ID", "Count", "Sample", "Lib_Type")
Enriched <- tmp %>% group_by(Sample) %>% mutate(Rel_ab = Count/sum(Count), Method = "Enriched") %>% merge(., GG_taxa, by = 1) %>% separate(Taxa, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = "; ", remove = F)  %>% ungroup
save(Enriched, file = "Enriched_feature_table.Rdata")
```
### Read Metagenomic sequencing data
Below is the unenriched library analyzed by MetaPhlan2
```{r eval=FALSE}
fn_path <- "Meta/2-Metaphlan"
fn <- list.files(fn_path, pattern = "_profile.txt", full.names = T)
fn_profile <- lapply(fn, function(x) {read.table(x, header = F, sep = "\t", stringsAsFactors = F, comment.char = "#")})
names(fn_profile) <- gsub("_profile.txt", "", basename(fn))

Unenriched <- lapply(seq_along(fn_profile), function(x){fn_profile[[x]] %>% mutate(Sample = names(fn_profile)[x])}) %>% do.call("rbind", .)
colnames(Unenriched) <- c("Taxa", "Rel_ab", "Coverage", "Avg_genome_length", "Count", "Sample")

Meta_unenriched <- Unenriched %>% mutate(Lib_Type = "PE150/Meta", Method = "Unenriched") %>% separate(Taxa, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Strain"), sep = "\\|", remove = F)  %>% ungroup
save(Meta_unenriched, file = "Meta_unenriched_feature_table.Rdata")
```
Below is the unenriched library analyzed by mapping to 16S database (the same method as the enriched library)
```{r eval=FALSE}
## metagenomic results based on bbmap to 16S database
Meta_16Smap <- read.table("Meta/Meta_PE150_OTU_count.txt", header = F, sep = "\t", stringsAsFactors = F)
colnames(Meta_16Smap) <- c("OTU_ID", "Count", "Sample", "Lib_Type")
Meta_16Smap$Lib_Type = "PE150/16SMap"
Meta_16Smap <- Meta_16Smap %>% mutate(Method = "Unenriched") %>% merge(., GG_taxa, by = 1) %>% separate(Taxa, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = "; ", remove = F) %>% ungroup
save(Meta_16Smap, file = "Meta_16Smap_feature_table.Rdata")
```

### Creating working dataset
```{r eval=FALSE}
load("Enriched_feature_table.Rdata") # Enriched
load("Meta_unenriched_feature_table.Rdata") # Meta_unenriched
load("Meta_16Smap_feature_table.Rdata") # Meta_16Smap
load("Amplicon/Amplicon_BG_feature_table.Rdata") # Amplicon_bg
load("Amplicon/Amplicon_V3V4_feature_table.Rdata") # Amplicon_v3v4

Amplicon_bg_int <- Amplicon_bg %>% gather(key = "SampleID", value = "Count", starts_with("BG")) %>% filter(Count>0) %>% group_by(SampleID) %>% mutate(Rel_ab = Count/sum(Count), Taxa = paste(Kingdom, Phylum, Class, Order, Family, Genus, Species, sep = "; "), SampleID = gsub("_F_filt.fastq.gz", "", SampleID), Lib_Type = "V4", Method = "Amplicon") %>% ungroup

Amplicon_v3v4_int <- Amplicon_v3v4 %>% gather(key = "SampleID", value = "Count", c(starts_with(c("JW", "Zymo")))) %>% filter(Count>0) %>% group_by(SampleID) %>% mutate(Rel_ab = Count/sum(Count), Taxa = paste(Kingdom, Phylum, Class, Order, Family, Genus, Species, sep = "; "), SampleID = gsub("_F_filt.fastq.gz", "", SampleID), Lib_Type = "V3V4", Method = "Amplicon") %>% ungroup

Amplicon_final <- rbind(Amplicon_bg_int, Amplicon_v3v4_int) %>% mutate(Sample = SampleID, SampleID = NULL)
#house keeping for Amplicon
Amplicon_final$Phylum[is.na(Amplicon_final$Phylum)] <- "p__"
Amplicon_final$Class[is.na(Amplicon_final$Class)] <- "c__"
Amplicon_final$Order[is.na(Amplicon_final$Order)] <- "o__"
Amplicon_final$Family[is.na(Amplicon_final$Family)] <- "f__"
Amplicon_final$Genus[is.na(Amplicon_final$Genus)] <- "g__"
Amplicon_final$Species[is.na(Amplicon_final$Species)] <- "s__"

Meta_unenriched_final <- Meta_unenriched %>% select(Sample, Rel_ab, Count, Taxa, c(2:9), Lib_Type, Method) %>% filter(!is.na(Species) & is.na(Strain)) %>% mutate(OTU_ID = NA) %>% select(-Strain) %>% ungroup

Meta_16Smap_final <- Meta_16Smap %>% group_by(Sample) %>% mutate(Rel_ab = Count/sum(Count)) %>% ungroup
```
```{r eval=FALSE}
coln <- intersect(colnames(Enriched), colnames(Amplicon_final)) %>% intersect(., colnames(Meta_unenriched_final)) %>% intersect(., colnames(Meta_16Smap_final))
Work <- rbind(Enriched %>% select(any_of(coln)), Amplicon_final %>% select(any_of(coln)), Meta_unenriched_final %>% select(any_of(coln)), Meta_16Smap_final %>% select(any_of(coln))) %>% filter(!grepl("^JT|JWHM277D", Sample))

## Unify the sample name
Work$Sample <- gsub("Bei-", "BG", Work$Sample, fixed = T)
Work$Sample[grepl("BEI-", Work$Sample)] <- "BEI.mock"
Work$Sample[grepl("Zymo", Work$Sample)] <- "Zymo.mock"
Work$Sample <- gsub("-S", "S", Work$Sample, fixed = T)
Work$Sample[Work$Sample=="JWHM276D"] <- "BEI.mock"

## house keeping
Work <- Work %>% mutate(Rel_ab = ifelse(Lib_Type=="PE150/Meta", Rel_ab/100, Rel_ab))
Work$Phylum[Work$Phylum=="p__[Thermi]"] <- "p__Deinococcus_Thermus" 

Mock_theory <- read.delim(file = "MockCommunities.txt", stringsAsFactors = F)
Mock_final <- Mock_theory %>% gather(key = "Sample", value = "Rel_ab", c(2:4), na.rm = T) %>% mutate(Taxa = paste(Kingdom, Phylum, Class, Order, Family, Genus, Species, sep = "; "))
Mock_final$Phylum[Mock_final$Phylum=="p__[Thermi]"] <- "p__Deinococcus_Thermus" 

saveRDS(Work, file = "Work_3MethodsCombined.Rds")
saveRDS(Mock_final, file = "Mock_communities.Rds")
```

## Analysis
Data files
```{r}
Work <- readRDS("Work_3MethodsCombined.Rds")
Mock_theory_final <- readRDS("Mock_communities.Rds")
```

### Mock community
```{r}
Mock_sample <- Work %>% filter(grepl("mock", Sample))

# create phylum level abundance graph
Mock_theory_p <- Mock_theory_final %>% filter(Sample!="JW.HM277D") %>% group_by(Method, Lib_Type, Sample, Kingdom, Phylum) %>% summarise(Rel_ab = sum(Rel_ab)) %>% ungroup

Mock_work_p <- Mock_sample %>% group_by(Method, Lib_Type, Sample, Kingdom, Phylum) %>% summarise(Rel_ab = sum(Rel_ab)) %>% 
    ungroup %>% merge(., Mock_theory_p[, 3:6], by = c("Sample", "Kingdom", "Phylum"), all.x = T) %>%  # merge with the thoeretical community composition
    mutate(Phylum2 = ifelse(is.na(Rel_ab.y), "z_Others", Phylum) %>% factor(levels = c("p__Actinobacteria", "p__Bacteroidetes", "p__Deinococcus_Thermus", "p__Firmicutes", "p__Proteobacteria", "z_Others"))) %>%
    rename(Rel_ab_s = Rel_ab.x, Rel_ab_t = Rel_ab.y) %>%
    mutate(Analysis_method = case_when(Method=="Amplicon" ~ "Denoising", Method=="Enriched" ~ "16S Mapping", grepl(pattern = "16S", Lib_Type) ~ "16S Mapping", grepl(pattern = "Meta", Lib_Type) ~ "Marker Gene"), Methods_cat = paste(Method, Analysis_method, sep = "\n"))

Mock_work_p2 <- Mock_work_p %>% filter(!is.na(Rel_ab_t)) %>% group_by(Sample, Phylum2) %>% mutate(Phylum3 = gsub("p__", "", Phylum2))
ggplot(Mock_work_p2, aes(x = Phylum3, y = Rel_ab_s*100, fill = Methods_cat)) +
    geom_crossbar(aes(ymin = Rel_ab_t*100, ymax = Rel_ab_t*100, y = Rel_ab_t*100), color = "black", width = 1, size = 0.2) +
    geom_point(size = 1.5, shape = 21, alpha = 0.7, position = position_jitter(width = 0.25))+
    #geom_beeswarm(size = 2, shape = 21, alpha = 0.5) +
    facet_wrap(.~Sample, scales = "free_y", ncol = 1) +
    coord_flip() +
    labs(x = "", y = "Relative abundance (%)", fill = "Methods") +
    theme_classic(base_size = 8) + theme(plot.title.position = "plot", panel.grid.major = element_line(color = "gray", size = 0.2))

set_panel_size(file = paste0("Figures/Lollipop_mock_phylum.pdf"), width = unit(3, "in"), height = unit(c(1.2, 0.5), "in"))


# create genus level abundance graph
Mock_theory_g <- Mock_theory_final %>% filter(Sample!="JW.HM277D") %>% group_by(Method, Lib_Type, Sample, Kingdom, Phylum, Class, Order, Family, Genus) %>% summarise(Rel_ab = sum(Rel_ab)) %>% ungroup

Mock_work_g <- Mock_sample %>% group_by(Method, Lib_Type, Sample, Kingdom, Phylum, Class, Order, Family, Genus) %>% summarise(Rel_ab = sum(Rel_ab)) %>% ungroup %>% split(., list(.$Method, .$Lib_Type), drop = T) %>% lapply(., function(x){merge(x, Mock_theory_g[, 3:10], by = c("Sample", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus"), all = T) %>% mutate(Method = Method[1], Lib_Type = Lib_Type[1])}) %>% do.call("rbind", .) %>%  # merge with the thoeretical community composition
    mutate(Genus2= ifelse(is.na(Rel_ab.y), "z_Others", Genus)) %>%
    rename(Rel_ab_s = Rel_ab.x, Rel_ab_t = Rel_ab.y) %>%
    mutate(Analysis_method = case_when(Method=="Amplicon" ~ "Denoising", Method=="Enriched" ~ "16S Mapping", grepl(pattern = "16S", Lib_Type) ~ "16S Mapping", grepl(pattern = "Meta", Lib_Type) ~ "Marker Gene"), Methods_cat = paste(Method, Analysis_method, sep = "\n"))

Mock_work_g2 <- Mock_work_g %>% filter(Family %in% c("f__Enterobacteriaceae", "f__Listeriaceae", "f__Bacillaceae"), Genus=="g__") %>% mutate(Genus2 = gsub("f__", "f_", Family))

Mock_work_g3 <- Mock_work_g %>% filter(!is.na(Rel_ab_t)) %>% rbind(., Mock_work_g2) %>% group_by(Sample, Genus2) %>% mutate(Genus3 = gsub("g__", "", Genus2))
Mock_work_g3$Genus3 <- factor(Mock_work_g3$Genus3, levels = Mock_work_g3 %>% ungroup() %>% arrange(Phylum, Class, Order, Family, Genus) %>% distinct(Phylum, Class, Order, Family, Genus, Genus3) %>% pull(Genus3))

Mock_work_g3 %>% ungroup() %>% distinct(Sample, Genus3) %>% select(Sample) %>% table

ggplot(Mock_work_g3, aes( x= Genus3, y = Rel_ab_s*100, fill = Methods_cat)) +
    geom_crossbar(aes(ymin = Rel_ab_t*100, ymax = Rel_ab_t*100, y = Rel_ab_t*100), color = "black", width = 1, size = 0.2) +
    geom_beeswarm(size = 1.5, shape = 21, alpha = 0.7) +
    facet_wrap(.~Sample, scales = "free_y", ncol = 1) +
    coord_flip() +
    labs(x = "", y = "Relative abundance (%)", fill = "Methods") +
    theme_classic(base_size = 8) + theme(plot.title.position = "plot", panel.grid.major = element_line(color = "gray", size = 0.2))
set_panel_size(file = paste0("Figures/Lollipop_mock_genus.pdf"), width = unit(3, "in"), height = unit(c(3, 1.5), "in"))

```

```{r}
# Check the accuracy of each sample
## use the dataset without corrections
Mock_g_s_fc <- Mock_work_g %>% filter(!is.na(Rel_ab_t)) %>% mutate(fc = ifelse(Rel_ab_s>Rel_ab_t, Rel_ab_s/Rel_ab_t, Rel_ab_t/Rel_ab_s)) %>% select(-Method, -Analysis_method, -Lib_Type)

ggplot(Mock_g_s_fc, aes(x = Methods_cat, y = fc)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(aes(color = Genus), position = position_jitter(width = 0.05)) +
    facet_grid(.~Sample) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 2) +
    labs(x = "Method, Sequencing length", y = "Fold Change (Upper or Under)\n between measured abundance vs nominal abundance", title = "Original comparison")
set_panel_size(file = "Figures/Boxplot_fc_g_org.pdf", width = unit(3, "in"), height = unit(6, "in"), panel.size = T, useDingbats = F)

## resolve the three taxa that is likely the miss in the original dataset
Mock_g_s_fc_edit1 <- Mock_work_g %>% filter(!is.na(Rel_ab_t), !is.na(Rel_ab_s))
Mock_g_s_fc_edit2 <- Mock_work_g %>% filter(!is.na(Rel_ab_t), is.na(Rel_ab_s)) %>% select(-Rel_ab_s) %>% merge(., Mock_work_g %>% filter(is.na(Rel_ab_t), Rel_ab_s>0.02) %>% select(-c(Genus, Rel_ab_t, Genus2)), all = T )
Mock_g_s_fc_edit <- rbind(Mock_g_s_fc_edit1, Mock_g_s_fc_edit2) %>% select(-Method, -Analysis_method, -Lib_Type)
Mock_g_s_fc_edit$Genus[is.na(Mock_g_s_fc_edit$Genus)] <- "g__Bacillus"
Mock_g_s_fc_edit$Genus2[is.na(Mock_g_s_fc_edit$Genus2)] <- "g__Bacillus"
Mock_g_s_fc_edit$Rel_ab_t[is.na(Mock_g_s_fc_edit$Rel_ab_t)] <- 0
Mock_g_s_fc_edit$Rel_ab_s[is.na(Mock_g_s_fc_edit$Rel_ab_s)] <- 0

Mock_g_s_fc_edit <- Mock_g_s_fc_edit %>% group_by(Methods_cat, Sample, Kingdom, Phylum, Class, Order, Family, Genus, Genus2) %>% summarise_all(., sum) %>% ungroup %>% mutate(fc = ifelse(Rel_ab_s>Rel_ab_t, Rel_ab_s/Rel_ab_t, Rel_ab_t/Rel_ab_s), genus_label = factor(Genus) %>% as.numeric())

Mock_g_s_fc_edit_duncan <- Mock_g_s_fc_edit %>% split(., .$Sample) %>% lapply(., function(x){(lm(fc ~ Methods_cat, data = x) %>% duncan.test(., trt = "Methods_cat"))$groups}) %>% do.call("rbind", .) %>% rownames_to_column(var = "UU") %>% separate(UU, into = c("Sample", "Methods_cat"), sep = "k\\.|-", ) %>% mutate(Sample = gsub("moc", "mock", Sample, fixed = T))

ggplot(Mock_g_s_fc_edit, aes(x = Methods_cat, y = fc)) +
    geom_boxplot(outlier.shape = NA) +
    geom_text(aes(label = genus_label, color = paste0(genus_label, ": ", Genus)), size = 8/.pt) +
    geom_text(data = Mock_g_s_fc_edit_duncan, aes(y = 0.8, label = groups), size = 7/.pt) +
    geom_text(aes(x = 2, y = 0.9, label = "Duncan's Multiple Range Test"), size = 7/.pt, hjust = 0) +
    facet_grid(.~Sample) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 3) +
    labs(x = "Method, Sequencing length", y = "Fold Change (Upper or Under)\n between measured abundance vs nominal abundance", color = "Genus", title = "With potential corrections")
set_panel_size(file = "Figures/Boxplot_fc_g_revised.pdf", width = unit(3.5, "in"), height = unit(6, "in"), panel.size = T, useDingbats = F)

Mock_g_s_fc_edit2 <- Mock_g_s_fc_edit %>% group_by(Sample, Genus2) %>% mutate(Genus3 = gsub("g__", "", Genus2))
Mock_g_s_fc_edit2$Genus3 <- factor(Mock_g_s_fc_edit2$Genus3, levels = Mock_work_g3 %>% ungroup() %>% arrange(Phylum, Class, Order, Family, Genus) %>% distinct(Phylum, Class, Order, Family, Genus, Genus3) %>% pull(Genus3))

ggplot(Mock_g_s_fc_edit2, aes(x= Genus3, y = fc, fill = Methods_cat)) +
    geom_beeswarm(size = 1.5, shape = 21, alpha = 0.7) +
    facet_wrap(.~Sample, scales = "free", ncol = 1) +
    coord_flip() +
    labs(x = "", y = "Fold Change (Upper or Under)\n between measured abundance vs nominal abundance", fill = "Methods") +
    theme_classic(base_size = 8) + theme(plot.title.position = "plot", panel.grid.major = element_line(color = "gray", size = 0.2))
set_panel_size(file = paste0("Figures/Lollipop_fc_g_revised.pdf"), width = unit(3, "in"), height = unit(c(3, 1.5), "in"))

```


### Diversity
#### Beta diversity
```{r}
# The diversity analysis was performed using family level composition
Work_f <- Work %>% group_by(across(c(Sample, Method, Lib_Type, Kingdom:Family))) %>% summarise(Rel_ab = sum(Rel_ab))
Work_f_ft <- Work_f %>% unite(col = SampleID, Sample:Lib_Type, sep = "_") %>% unite(col = Taxa, Kingdom:Family, sep = ";") %>% mutate(Taxa = gsub("__", "_", Taxa, fixed = T)) %>% spread(key = Taxa, value = Rel_ab, fill = 0) %>% column_to_rownames(var = "SampleID")

# Bray_curtis
Family_bray <- vegdist(Work_f_ft, method = "bray")
Family_bray_pcoa <- pcoa(Family_bray)

Family_bray_data <- Family_bray_pcoa$vectors[,1:5] %>% as.data.frame() %>% rownames_to_column(var = "ID") %>% separate(col = ID, into = c("Sample", "Library"), sep = "_", remove = F, extra = "merge") %>% mutate(Projects = ifelse(grepl("mock", Sample), Sample, ifelse(grepl("JWS", Sample), "Rat", "Mouse"))) %>% mutate(SampleNo = factor(Sample) %>% as.numeric)

varPC1=round(Family_bray_pcoa$values$Relative_eig[1]*100, 2)
varPC2=round(Family_bray_pcoa$values$Relative_eig[2]*100, 2)
varPC3=round(Family_bray_pcoa$values$Relative_eig[3]*100, 2)
varPC4=round(Family_bray_pcoa$values$Relative_eig[4]*100, 2)

ggplot(Family_bray_data, aes(x = Axis.1, y = Axis.2, shape = Library, fill = Projects)) +
    geom_point(size = 2) +
    geom_text_repel(aes(label = SampleNo), size = 6/.pt, color = "black", segment.size = 0.2) + 
    scale_shape_manual(values = c(21, 21, 22, 23, 24, 25))+
    scale_color_manual(values = c('#e41a1c','#377eb8','#4daf4a','#984ea3')) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 1, panel.border = element_rect(color = "black", fill = NA), panel.grid.major = element_line(color = "grey90", linetype = "dashed", size = 0.2)) +
    labs(x=paste0("PC1: ", varPC1, "% explained variation"), y=paste0("PC2: ", varPC2, "% explained variation"), title = "PCoA analysis on Bray-Curtis dissimilarity \nbased on family level taxa profile", subtitle = "PC1 vs PC2") +
    guides(fill= guide_legend(override.aes = list(size = 2.5, shape = 21)))
set_panel_size(file = "Figures/Beta_Bray_family_PC1vsPC2.pdf", width = unit(3.5, "in"), height = unit(3.5, "in"), useDingbats = F)

# Jaccard
Family_jacc <- vegdist(Work_f_ft, method = "jaccard", binary = T)
Family_jacc_pcoa <- pcoa(Family_jacc)

Family_jacc_data <- Family_jacc_pcoa$vectors[,1:5] %>% as.data.frame() %>% rownames_to_column(var = "ID") %>% separate(col = ID, into = c("Sample", "Library"), sep = "_", remove = F, extra = "merge") %>% mutate(Projects = ifelse(grepl("mock", Sample), Sample, ifelse(grepl("JWS", Sample), "Rat", "Mouse"))) %>% mutate(SampleNo = factor(Sample) %>% as.numeric)

varPC1=round(Family_jacc_pcoa$values$Relative_eig[1]*100, 2)
varPC2=round(Family_jacc_pcoa$values$Relative_eig[2]*100, 2)
varPC3=round(Family_jacc_pcoa$values$Relative_eig[3]*100, 2)
varPC4=round(Family_jacc_pcoa$values$Relative_eig[4]*100, 2)

ggplot(Family_jacc_data, aes(x = Axis.1, y = Axis.2, shape = Library, fill = Projects)) +
    geom_point(size = 2) +
    geom_text_repel(aes(label = SampleNo), size = 6/.pt, color = "black", segment.size = 0.2) + 
    scale_shape_manual(values = c(21, 21, 22, 23, 24, 25))+
    scale_color_manual(values = c('#e41a1c','#377eb8','#4daf4a','#984ea3')) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 1, panel.border = element_rect(color = "black", fill = NA), panel.grid.major = element_line(color = "grey90", linetype = "dashed", size = 0.2)) +
    labs(x=paste0("PC1: ", varPC1, "% explained variation"), y=paste0("PC2: ", varPC2, "% explained variation"), title = "PCoA analysis on Jaccard dissimilarity \nbased on family level taxa profile", subtitle = "PC1 vs PC2") +
    guides(fill= guide_legend(override.aes = list(size = 2.5, shape = 21)))
set_panel_size(file = "Figures/Beta_jacc_family_PC1vsPC2.pdf", width = unit(3.5, "in"), height = unit(3.5, "in"), useDingbats = F)


```

```{r}
# The diversity analysis was performed using Genus level composition
Work_g <- Work %>% group_by(across(c(Sample, Method, Lib_Type, Kingdom:Genus))) %>% summarise(Rel_ab = sum(Rel_ab))
Work_g_ft <- Work_g %>% unite(col = SampleID, Sample:Lib_Type, sep = "_") %>% unite(col = Taxa, Kingdom:Genus, sep = ";") %>% mutate(Taxa = gsub("__", "_", Taxa, fixed = T)) %>% spread(key = Taxa, value = Rel_ab, fill = 0) %>% column_to_rownames(var = "SampleID")

# Bray_curtis
Genus_bray <- vegdist(Work_g_ft, method = "bray")
Genus_bray_pcoa <- pcoa(Genus_bray)

Genus_bray_data <- Genus_bray_pcoa$vectors[,1:5] %>% as.data.frame() %>% rownames_to_column(var = "ID") %>% separate(col = ID, into = c("Sample", "Library"), sep = "_", remove = F, extra = "merge") %>% mutate(Projects = ifelse(grepl("mock", Sample), Sample, ifelse(grepl("JWS", Sample), "Rat", "Mouse"))) %>% mutate(SampleNo = factor(Sample) %>% as.numeric)

varPC1=round(Genus_bray_pcoa$values$Relative_eig[1]*100, 2)
varPC2=round(Genus_bray_pcoa$values$Relative_eig[2]*100, 2)
varPC3=round(Genus_bray_pcoa$values$Relative_eig[3]*100, 2)
varPC4=round(Genus_bray_pcoa$values$Relative_eig[4]*100, 2)

ggplot(Genus_bray_data, aes(x = Axis.1, y = Axis.2, shape = Library, fill = Projects)) +
    geom_point(size = 2) +
    geom_text_repel(aes(label = SampleNo), size = 6/.pt, color = "black", segment.size = 0.2) + 
    scale_shape_manual(values = c(21, 21, 22, 23, 24, 25))+
    scale_color_manual(values = c('#e41a1c','#377eb8','#4daf4a','#984ea3')) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 1, panel.border = element_rect(color = "black", fill = NA), panel.grid.major = element_line(color = "grey90", linetype = "dashed", size = 0.2)) +
    labs(x=paste0("PC1: ", varPC1, "% explained variation"), y=paste0("PC2: ", varPC2, "% explained variation"), title = "PCoA analysis on Bray-Curtis dissimilarity \nbased on Genus level taxa profile", subtitle = "PC1 vs PC2") +
    guides(fill= guide_legend(override.aes = list(size = 2.5, shape = 21)))
set_panel_size(file = "Figures/Beta_Bray_Genus_PC1vsPC2.pdf", width = unit(3.5, "in"), height = unit(3.5, "in"), useDingbats = F)

# jaccard
Genus_jacc <- vegdist(Work_g_ft, method = "jaccard", binary = T)
Genus_jacc_pcoa <- pcoa(Genus_jacc)

Genus_jacc_data <- Genus_jacc_pcoa$vectors[,1:5] %>% as.data.frame() %>% rownames_to_column(var = "ID") %>% separate(col = ID, into = c("Sample", "Library"), sep = "_", remove = F, extra = "merge") %>% mutate(Projects = ifelse(grepl("mock", Sample), Sample, ifelse(grepl("JWS", Sample), "Rat", "Mouse"))) %>% mutate(SampleNo = factor(Sample) %>% as.numeric)

varPC1=round(Genus_jacc_pcoa$values$Relative_eig[1]*100, 2)
varPC2=round(Genus_jacc_pcoa$values$Relative_eig[2]*100, 2)
varPC3=round(Genus_jacc_pcoa$values$Relative_eig[3]*100, 2)
varPC4=round(Genus_jacc_pcoa$values$Relative_eig[4]*100, 2)

ggplot(Genus_jacc_data, aes(x = Axis.1, y = Axis.2, shape = Library, fill = Projects)) +
    geom_point(size = 2) +
    geom_text_repel(aes(label = SampleNo), size = 6/.pt, color = "black", segment.size = 0.2) + 
    scale_shape_manual(values = c(21, 21, 22, 23, 24, 25))+
    scale_color_manual(values = c('#e41a1c','#377eb8','#4daf4a','#984ea3')) +
    theme_classic(base_size = 10) + theme(aspect.ratio = 1, panel.border = element_rect(color = "black", fill = NA), panel.grid.major = element_line(color = "grey90", linetype = "dashed", size = 0.2)) +
    labs(x=paste0("PC1: ", varPC1, "% explained variation"), y=paste0("PC2: ", varPC2, "% explained variation"), title = "PCoA analysis on Jaccard dissimilarity \nbased on Genus level taxa profile", subtitle = "PC1 vs PC2") +
    guides(fill= guide_legend(override.aes = list(size = 2.5, shape = 21)))
set_panel_size(file = "Figures/Beta_jacc_Genus_PC1vsPC2.pdf", width = unit(3.5, "in"), height = unit(3.5, "in"), useDingbats = F)
```

#### Distance
```{r}
Family_bray_dist <- tidy(Family_bray)
Family_bray_dist_tbl <- Family_bray_dist %>% separate(item1, into = c("Sample1", "Library1"), sep = "_", extra = "merge") %>% separate(item2, into = c("Sample2", "Library2"), sep = "_", extra = "merge") %>% filter(Sample1==Sample2) %>% mutate(Projects = ifelse(grepl("mock", Sample1), "Mock", ifelse(grepl("JWS", Sample1), "Rat", "Mouse")), Comparison = paste(Library1, Library2, sep = "v"))

Family_bray_dist_tbl_duncan <- Family_bray_dist_tbl %>% filter(Projects!="Mock") %>% split(., .$Projects) %>% lapply(., function(x){(lm(distance ~ Comparison, data = x) %>% duncan.test(., trt = "Comparison"))$groups}) %>% do.call("rbind", .) %>% rownames_to_column(var = "UU") %>% separate(UU, into = c("Projects", "Comparison"), sep = "\\.", )

ggplot(Family_bray_dist_tbl %>% filter(Projects!="Mock"), aes(x = Comparison, y = distance)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 1, position = position_jitter(width = 0.1))+
    facet_grid(.~Projects, scales = "free_x") +
    geom_text(data = Family_bray_dist_tbl_duncan, aes(y = -0.05, label = groups), size = 6/.pt) +
    geom_text(aes(x = 2, y = 0, label = "Duncan's Multiple Range Test"), size = 6/.pt, hjust = 0) +
    theme_classic(base_size = 8) + theme(axis.text.x = element_text(angle = 90, hjust =1)) +
    labs(x = "", y = "Bray-Curtis Distance", title = "Comparison of bray-curtis distance", subtitle = "Based on Family level composition")
set_panel_size(file = "Figures/Bray_distance_family.pdf", width = unit(3, "in"), height = unit(3, "in"), useDingbats = F)

Genus_bray_dist <- tidy(Genus_bray)
Genus_bray_dist_tbl <- Genus_bray_dist %>% separate(item1, into = c("Sample1", "Library1"), sep = "_", extra = "merge") %>% separate(item2, into = c("Sample2", "Library2"), sep = "_", extra = "merge") %>% filter(Sample1==Sample2) %>% mutate(Projects = ifelse(grepl("mock", Sample1), "Mock", ifelse(grepl("JWS", Sample1), "Rat", "Mouse")), Comparison = paste(Library1, Library2, sep = "v"))

Genus_bray_dist_tbl_duncan <- Genus_bray_dist_tbl %>% filter(Projects!="Mock") %>% split(., .$Projects) %>% lapply(., function(x){(lm(distance ~ Comparison, data = x) %>% duncan.test(., trt = "Comparison"))$groups}) %>% do.call("rbind", .) %>% rownames_to_column(var = "UU") %>% separate(UU, into = c("Projects", "Comparison"), sep = "\\.", )

ggplot(Genus_bray_dist_tbl %>% filter(Projects!="Mock"), aes(x = Comparison, y = distance)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 1, position = position_jitter(width = 0.1))+
    facet_grid(.~Projects, scales = "free_x") +
    geom_text(data = Genus_bray_dist_tbl_duncan, aes(y = -0.05, label = groups), size = 6/.pt) +
    geom_text(aes(x = 2, y = 0, label = "Duncan's Multiple Range Test"), size = 6/.pt, hjust = 0) +
    theme_classic(base_size = 8) + theme(axis.text.x = element_text(angle = 90, hjust =1)) +
    labs(x = "", y = "Bray-Curtis Distance", title = "Comparison of bray-curtis distance", subtitle = "Based on Genus level composition")
set_panel_size(file = "Figures/Bray_distance_Genus.pdf", width = unit(3, "in"), height = unit(3, "in"), useDingbats = F)


```



### Rarefaction curve

```{r}
sample_dict = "Bei-1	MF1
Bei-2	MF2
Bei-3	MF3
Bei-4	MF4
Bei-11	MF5
Bei-12	MF6
Bei-13	MF7
Bei-14	MF8
JW-S09	RF1
JW-S12	RF2
JW-S14	RF3
JW-S18	RF4
JW-S22	RF5
" %>% read.table(text = ., header = F, sep = "\t", col.names = c("O_sample", "Sample"))

Enriched150 <- read.table("16Scap/Enriched_PE150_OTU_count.txt", header = F, sep = "\t", stringsAsFactors = F) # a table with OTU id, count, sampleid, and library type
colnames(Enriched150) <- c("OTU_ID", "Count", "Sample", "Lib_Type")

Enriched_ft <- Enriched150 %>% filter(!grepl("JT-|mock", Sample)) %>% pivot_wider(id_cols =OTU_ID, names_from = Sample, values_from = Count, values_fill = 0) %>% column_to_rownames(var = "OTU_ID")
Enriched_ft_otu_ct = colSums(Enriched_ft)

Enriched_rare_nspec = rarefy(Enriched_ft, c(seq(2E2, 5E3, 2E2), seq(1E4, 4.5E5, 2E4)), MARGIN = 2)

Enriched_rare_nspec_tb <- Enriched_rare_nspec %>% as.tibble() %>% mutate(Sample = colnames(Enriched_ft)) %>% pivot_longer(cols = starts_with("N"), names_to = "Depth", values_to = "nspec") %>% mutate(Depth = as.numeric(gsub("N", "", Depth)), Method = "Enriched", TotalOTUCt = Enriched_ft_otu_ct[Sample]) %>% filter(Depth<= TotalOTUCt)


Meta_16Smap <- read.table("Meta/Meta_PE150_OTU_count.txt", header = F, sep = "\t", stringsAsFactors = F)
colnames(Meta_16Smap) <- c("OTU_ID", "Count", "Sample", "Lib_Type")

Meta_ft <- Meta_16Smap %>% filter(!grepl("JT-|mock", Sample)) %>% pivot_wider(id_cols =OTU_ID, names_from = Sample, values_from = Count, values_fill = 0) %>% column_to_rownames(var = "OTU_ID")
Meta_ft_otu_ct = colSums(Meta_ft)

Meta_rare_nspec = rarefy(Meta_ft, seq(2E2, 4E3, 2E2), MARGIN = 2)
Meta_rare_nspec_tb <- Meta_rare_nspec %>% as.tibble() %>% mutate(Sample = colnames(Meta_ft)) %>% pivot_longer(cols = starts_with("N"), names_to = "Depth", values_to = "nspec") %>% mutate(Depth = as.numeric(gsub("N", "", Depth)), Method = "Unenriched", TotalOTUCt = Meta_ft_otu_ct[Sample]) %>% filter(Depth<= TotalOTUCt)

Rare_work <- rbind(Enriched_rare_nspec_tb, Meta_rare_nspec_tb) %>% merge(sample_dict, ., by = 1) %>% select(-O_sample) %>% mutate(Project = ifelse(grepl("MF", Sample), "Mouse Feces", "Rat Feces"))

p1 = ggplot(Rare_work[Rare_work$Project=="Mouse Feces", ], aes(x = Depth, y = nspec, color = Sample)) +
    geom_line(aes(linetype = Method)) +
    theme_classic(base_size = 8) + theme(aspect.ratio = 0.8) +
    scale_color_manual(values =  c('#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00','#ffff33','#a65628','#f781bf')) +
    scale_x_continuous(labels = scales::comma) +
    guides(color = F) +
    labs(x = "Sequencing depth", y = "Number of OTUs")
print(p1)

p2 = ggplot(Rare_work[Rare_work$Project=="Mouse Feces" & Rare_work$Depth<=3000, ], aes(x = Depth, y = nspec, color = Sample)) +
    geom_line(aes(linetype = Method)) +
    theme_classic(base_size = 8) + theme(aspect.ratio = 0.8, legend.position = "none") +
    scale_color_manual(values =  c('#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00','#ffff33','#a65628','#f781bf')) +
    scale_x_continuous(labels = scales::comma) +
    labs(x = "Sequencing depth", y = "Number of OTUs")
print(p2)

p3 = p1 + inset_element(p2, left = 0.5, bottom = 0, top = 0.6, right = 1)
print(p3)

p4 = ggplot(Rare_work[Rare_work$Project=="Rat Feces", ], aes(x = Depth, y = nspec, color = Sample)) +
    geom_line(aes(linetype = Method)) +
    theme_classic(base_size = 8) + theme(aspect.ratio = 0.8) +
    scale_color_manual(values =  c('#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00')) +
    scale_x_continuous(labels = scales::comma) +
    guides(color = F) +
    labs(x = "Sequencing depth", y = "Number of OTUs")
print(p4)

p5 = ggplot(Rare_work[Rare_work$Project=="Rat Feces" & Rare_work$Depth<=3000, ], aes(x = Depth, y = nspec, color = Sample)) +
    geom_line(aes(linetype = Method)) +
    theme_classic(base_size = 8) + theme(aspect.ratio = 0.8, legend.position = "none") +
    scale_color_manual(values =  c('#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00')) +
    scale_x_continuous(labels = scales::comma) +
    labs(x = "Sequencing depth", y = "Number of OTUs")
print(p5)

p6 = p4 + inset_element(p5, left = 0.5, bottom = 0, top = 0.6, right = 1)
print(p6)

p3 + p6 + plot_layout(guides = "collect", ncol = 1)
ggsave("Figures/Rare_no_filter.pdf", width = 5, useDingbats = F)
```

